---
title: "Reproducible Research - Course Project 1"
output:
  html_document: 
    keep_md: yes
---

```{r}
library(knitr)
opts_chunk$set(dev="png")
```

## Loading and preprocessing the data

**1. Loading the data**

```{r}
if (!file.exists('activity.csv')) {
    unzip(zipfile = "activity.zip")
}
activitydata <- read.csv("activity.csv")
```

**2. Transforming data as required**
```{r}
activitydata$date <- as.POSIXct(activitydata$date, format = "%Y-%m-%d")
```

## Histogram of total number of steps taken each day.

```{r total number of steps taken each day}
stepsperday <- aggregate(steps ~ date, activitydata, FUN = sum)
hist(stepsperday$steps, col = "gray", xlab = "Number of steps", ylab = "Frequency", main = "Total number of steps taken per day")
stepsmean <- mean(stepsperday$steps, na.rm = TRUE)
stepsmedian <- median(stepsperday$steps, na.rm = TRUE)
```

**Mean** of the total number of steps taken per day: **`r stepsmean`.**

**Median** of the total number of steps taken per day: **`r stepsmedian`.**

## Time series plot

```{r Time series plot}
avgstepsperday <- aggregate(steps ~ interval, activitydata, FUN = mean)
plot(avgstepsperday$interval, avgstepsperday$steps, type = "l", col = "Brown", lwd = 2, xlab = "5-minute intervals", ylab = "Average number of steps", main = "Average daily activity pattern")
maxstepsinterval <- avgstepsperday$interval[which.max(avgstepsperday$steps)]
```

On an average most steps were taken at interval **`r maxstepsinterval`**


## Imputing missing values

```{r}
totalNAs <- sum(is.na(activitydata))
```

Total number of missing values in the dataset is **`r totalNAs`**

**Imputed Data**

```{r}
imp_activitydata <- activitydata
imp_activitydata$steps[is.na(imp_activitydata$steps)] = avgstepsperday$steps[match(avgstepsperday$interval, imp_activitydata$interval)]
head(imp_activitydata)
```

```{r Total number of steps taken per day (Imputed data)}
imp_stepsperday <- aggregate(steps ~ date, imp_activitydata, FUN = sum)
hist(imp_stepsperday$steps, col = "purple", xlab = "Number of steps", ylab = "Frequency", main = "Total number of steps taken per day (Imputed data)")
imp_stepsmean <- mean(imp_stepsperday$steps, na.rm = TRUE)
imp_stepsmedian <- median(imp_stepsperday$steps, na.rm = TRUE)
```

**Mean** of the total number of steps taken per day (Imputed data): **`r imp_stepsmean`.**

**Median** of the total number of steps taken per day (Imputed data): **`r imp_stepsmedian`.**


**The only change observed between the two histograms for raw data and imputed data is the increase in frequency where 10000 to 15000 steps occured.**


## Differences in activity patterns between weekdays and weekends

```{r patterns between weekdays and weekends}
library(dplyr)
library(lattice)
week_activitydata <- imp_activitydata %>%
    mutate(day = weekdays(date)) %>%
    mutate(daytype = ifelse(day %in% c("Saturday","Sunday"), "weekend", "weekday"))
week_stepsperday <- aggregate(steps ~ interval + daytype, week_activitydata, FUN = mean)
xyplot(steps ~ interval | daytype, week_stepsperday, type="l", lwd=2, xlab="Interval", ylab="Number of steps", layout=c(1,2))
```
